CF-SuGAR Power Calculations

### 2025-03-18 Christine Chan/CF-SuGAR/power_calculations.qmd
NIH power calculations with Laura's original code

```{r setup}
#| include: false
library(Hmisc)
library(tidyverse)
library(childsds)
library(survival)
library(gtsummary)
library(ggstats)
home_dir <- switch(Sys.info()["sysname"],
  "Darwin" = "/Users/timvigers/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF/Christine Chan/CF Registry",
  "Linux" = "/home/timvigers/OneDrive/Vigers/CF/Christine Chan/CF Registry"
)
github_dir <- switch(Sys.info()["sysname"],
  "Darwin" = "/Users/timvigers/Documents/GitHub",
  "Linux" = "/home/timvigers/Documents/GitHub"
)
knitr::opts_knit$set(root.dir = home_dir)
```

@@

annual <- read.csv("./Data_Raw/CFF19_Annualized_Del1.csv",

@@

data_dict <- readxl::read_excel("./Data_Raw/Copy of Codes for CFFPR_2018.xlsx")

@@

# Get mutations
demogs <- read.csv("./Data_Raw/CFF19_DemogCFDiag_Del1.csv")

@@

# List of covariates
vars <- c(
  "age_group", "Gender", "bmi_perc", "bmi_cat", "MutClass", "Race",
  "Hispanicrace", "A_IsOnEnzymes", "ppFEV1_cat", "tube_feeding",
  "NumPulmExacerbation", "A_pseudomonasaeruginosa", "A_fungalyeast1",
  "A_mycocultureresults3", "A_corticosteroids1", "cfrd_status_annual",
  "A_Vx770", "A_VX809comb", "liver_disease"

@@

```{r}
# Origin for survival analyses
og <- ymd("2003-01-01")
# Need to use a dataframe not a tibble
annual <- as.data.frame(annual)
annual$tstart <- as.numeric(annual$tstart) - as.numeric(og)
annual$tstop <- as.numeric(annual$tstop) - as.numeric(og)
# Formula for base model
xvars <- c(
  "age_group", "Gender", "Race", "Hispanicrace", "bmi_cat", "MutClass",
  "A_IsOnEnzymes", "tube_feeding", "liver_disease", "ppFEV1_cat",
  "NumPulmExacerbation", "A_corticosteroids1"
)
y <- "Surv(time = tstart, time2 = tstop, event = as.numeric(CFRD) == 2)"
f <- as.formula(paste0(y, "~", paste0(xvars, collapse = "+")))
cox_res <- coxph(f, data = annual, cluster = eDWID)
# Forest plot and coefficients table
cox_res %>%
  tbl_regression(exponentiate = T)
ggcoef_table(cox_res,
  show_p_values = F, add_reference_rows = T, exponentiate = T,
  signif_stars = F, colour = F
)
```

@@

@@

ogtt <- read.csv("./Christine Chan/1 hour OGTT in CFRD/Data_raw/ogtt.csv")
pft <- read.csv("./Christine Chan/1 hour OGTT in CFRD/Data_raw/pfts.csv")
a1c <- read.csv("./Christine Chan/1 hour OGTT in CFRD/Data_raw/a1c.csv")
demo <- read.csv("./Christine Chan/1 hour OGTT in CFRD/Data_raw/demographics.csv")
dob <- read.csv("./Christine Chan/1 hour OGTT in CFRD/Data_raw/dob.csv")

@@

# read in date of modulator use
mod <- read.csv("./Christine Chan/1 hour OGTT in CFRD/Data_raw/modulators.csv", na.strings = c("", " ", NA))
mod$Kalydeco.Start.Date <- as.Date(mod$Kalydeco.Start.Date, format = "%m/%d/%Y")
mod$Orkambi.Start.Date <- as.Date(mod$Orkambi.Start.Date, format = "%m/%d/%Y")
mod$Symdeko.Start.Date <- as.Date(mod$Symdeko.Start.Date, format = "%m/%d/%Y")
mod$Trikafta.Start.Date <- as.Date(mod$Trikafta.Start.Date, format = "%m/%d/%Y")
mod$modstart <- as.Date(NA)
mod$he_modstart <- as.Date(NA)
mod$me_modstart <- as.Date(NA)
for (i in 1:nrow(mod)) {
  mod[i, ]$modstart <-
    as.Date(pmin(mod[i, ]$Kalydeco.Start.Date, mod[i, ]$Orkambi.Start.Date, mod[i, ]$Symdeko.Start.Date, mod[i, ]$Trikafta.Start.Date, na.rm = T))
  mod[i, ]$he_modstart <- as.Date(pmin(mod[i, ]$Kalydeco.Start.Date, mod[i, ]$Trikafta.Start.Date, na.rm = T))
  mod[i, ]$me_modstart <- as.Date(pmin(mod[i, ]$Orkambi.Start.Date, mod[i, ]$Symdeko.Start.Date, na.rm = T))
}
modkeep <- mod[, c("MRN", "modstart", "he_modstart", "me_modstart")]
alldata <- merge(alldata, modkeep, by = "MRN", all.x = TRUE, all.y = FALSE)
alldata$onmod <- ifelse(!is.na(alldata$modstart) & alldata$modstart <= alldata$Date, 1,
  ifelse(!is.na(alldata$modstart) & alldata$modstart > alldata$Date, 0, NA)
)
alldata$onmod <- ifelse(is.na(alldata$modstart) & is.na(alldata$onmod), 0, alldata$onmod)

@@

##############
# DESC STATS #
##############

# table of demographics in those with at least one pre-CFRD OGTT
demovars <- c(
  "Sex", "race_eth", "age", "genoRisk", "Fasting.Glucose", "X1.hour.Glucose", "X2.hour.Glucose", "onmod_baseline", "modstarted_fup",
  "A1C", "FEV1.per.pred", "FVC.per.pred", "BMI", "bmipercentile", "FEV.1.per.pred_FVC.per.pred"
)
# Taking out these clinical variables - if they are to be added to table 1, need to figure out how
# because they come from PFT visits, not the same date as OGTT
#              "BMI","FEV1.per.pred","FVC.per.pred","FEF.25.75.per.pred","bmip")
t1_alldata <- CreateTableOne(data = baseline_all, vars = demovars)
t1_alldata <- print(t1_alldata, printToggle = F, varLabels = TRUE, missing = T, showAllLevels = T, test = F)

# table of demographics in those with at least one pre-CFRD OGTT by sex
t1_alldata_sex <- CreateTableOne(data = baseline_all, vars = demovars, strata = "Sex")
t1_alldata_sex <- print(t1_alldata_sex, printToggle = F, varLabels = TRUE, missing = T, showAllLevels = T)

# table of demographics in those with more than 1 OGTT
t1_multogtt <- CreateTableOne(data = baseline_mult, vars = c(demovars, "Follow_up_time_yrs"))
t1_multogtt <- print(t1_multogtt, printToggle = F, varLabels = TRUE, missing = T, showAllLevels = T, test = F)

# table of demographics in those with more than 1 OGTT stratified by progression
t1_multogtt_prog <- CreateTableOne(data = baseline_mult, vars = c(demovars, "Follow_up_time_yrs", "age_at_CFRD_dx"), strata = "CFRD")
t1_multogtt_prog <- print(t1_multogtt_prog, printToggle = F, varLabels = TRUE, missing = T, showAllLevels = T, test = T)

# table of demographics in those with more than 1 OGTT by sex
t1_multogtt_sex <- CreateTableOne(data = baseline_mult, vars = demovars, strata = "Sex")
t1_multogtt_sex <- print(t1_multogtt_sex, printToggle = F, varLabels = TRUE, missing = T, showAllLevels = T)

# compare baseline glucose values in those who developed CFRD and those who did not
# added other vars too
gluctable <- CreateTableOne(baseline_mult, vars = demovars, strata = "CFRD")
gluctable <- print(gluctable, printToggle = F, varLabels = TRUE, missing = T, showAllLevels = T, test = T)

########################
# LOGISTIC REGRESSIONS #
########################

# logistic regression of 1 hour and 2 hour glucose predicting cfrd
log1 <- glm(data = baseline_mult, CFRD ~ Baseline_1hr_glucose, family = "binomial")
or1 <- exp(cbind(OR = coef(log1), confint(log1)))
log1_adj <- glm(data = baseline_mult, CFRD ~ Baseline_1hr_glucose + Base_age + Sex, family = "binomial")
or1_adj <- exp(cbind(OR = coef(log1_adj), confint(log1_adj)))

# add interaction with sex
log1_adj_int <- glm(data = baseline_mult, CFRD ~ Baseline_1hr_glucose + Base_age + Sex + Sex * Baseline_1hr_glucose, family = "binomial")
or1_adj_int <- exp(cbind(OR = coef(log1_adj_int), confint(log1_adj_int)))

log1_cat <- glm(data = baseline_mult, CFRD ~ Baseline_1hr_glucose_155, family = "binomial")
or1_cat <- exp(cbind(OR = coef(log1_cat), confint(log1_cat)))
log1_cat_adj <- glm(data = baseline_mult, CFRD ~ Baseline_1hr_glucose_155 + Base_age + Sex, family = "binomial")
or1_cat_adj <- exp(cbind(OR = coef(log1_cat_adj), confint(log1_cat_adj)))

log1_cat_140 <- glm(data = baseline_mult, CFRD ~ Baseline_1hr_glucose_140, family = "binomial")
or1_cat_140 <- exp(cbind(OR = coef(log1_cat_140), confint(log1_cat_140)))
log1_cat_adj_140 <- glm(data = baseline_mult, CFRD ~ Baseline_1hr_glucose_140 + Base_age + Sex, family = "binomial")
or1_cat_adj_140 <- exp(cbind(OR = coef(log1_cat_adj_140), confint(log1_cat_adj_140)))

# add interaction with sex
log1_cat_adj_int <- glm(
  data = baseline_mult, CFRD ~ Baseline_1hr_glucose_155 + Base_age + Sex + Sex * Baseline_1hr_glucose_155,
  family = "binomial"
)
or1_cat_adj_int <- exp(cbind(OR = coef(log1_cat_adj_int), confint(log1_cat_adj_int)))

log2 <- glm(data = baseline_mult, CFRD ~ Baseline_2hr_glucose, family = "binomial")
or2 <- exp(cbind(OR = coef(log2), confint(log2)))

log2_adj <- glm(data = baseline_mult, CFRD ~ Baseline_2hr_glucose + Base_age + Sex, family = "binomial")
or2_adj <- exp(cbind(OR = coef(log2_adj), confint(log2_adj)))

# add interaction with sex
log2_adj_int <- glm(data = baseline_mult, CFRD ~ Baseline_2hr_glucose + Base_age + Sex + Sex * Baseline_2hr_glucose, family = "binomial")
or2_adj_int <- exp(cbind(OR = coef(log2_adj_int), confint(log2_adj_int)))

# ROC for 1 hour glucose
# get predicted values from logistic regression
baseline_mult$pred_1hr <- predict(log1, type = "response", data.frame = baseline_mult)
# perform the ROC analysis
roc1 <- roc(data = baseline_mult, response = CFRD, predictor = X1.hour.Glucose)
# find the cutoff that maximizes the Youden Index
coords1 <- coords(roc1, "b", ret = c("t", "specificity", "sensitivity"), best.method = "youden") # default

# ROC for 2 hour glucose
baseline_mult$pred_2hr <- predict(log2, type = "response", data.frame = baseline_mult)
roc2 <- roc(data = baseline_mult, response = CFRD, predictor = X2.hour.Glucose)
coords2 <- coords(roc2, "b", ret = c("t", "specificity", "sensitivity"), best.method = "youden") # default

##############
# COX MODELS #
##############

# CPH model for baseline 1 hour glucose
cox1hr <- coxph(Surv(Follow_up_time_yrs, CFRD) ~ Baseline_1hr_glucose, data = baseline_mult)
```

>>>>

```{r setup}
#| include: false
library(Hmisc)
library(tidyverse)
library(childsds)
library(rspiro)
library(survival)
library(gtsummary)
library(ggstats)
home_dir <- switch(Sys.info()["sysname"],
  "Darwin" = "/Users/timvigers/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF/Christine Chan",
  "Linux" = "/home/timvigers/OneDrive/Vigers/CF/Christine Chan"
)
github_dir <- switch(Sys.info()["sysname"],
  "Darwin" = "/Users/timvigers/Documents/GitHub",
  "Linux" = "/home/timvigers/Documents/GitHub"
)
knitr::opts_knit$set(root.dir = home_dir)
```

@@

annual <- read.csv("./CF Registry/Data_Raw/CFF19_Annualized_Del1.csv",

@@

data_dict <- readxl::read_excel("./CF Registry/Data_Raw/Copy of Codes for CFFPR_2018.xlsx")

@@

# Get mutations, race, etc.
demogs <- read.csv("./CF Registry/Data_Raw/CFF19_DemogCFDiag_Del1.csv", na.strings = "")

@@

@@

```{r}
# Origin for survival analyses
og <- ymd("2003-01-01")
# Need to use a dataframe not a tibble
annual <- as.data.frame(annual)
annual$tstart <- as.numeric(annual$tstart) - as.numeric(og)
annual$tstop <- as.numeric(annual$tstop) - as.numeric(og)
# Formula for base model
xvars <- c(
  "Age_YrEnd", "Gender", "bmi", "MutClass", "tube_feeding", "liver_disease", "ppFEV1_cat",
  "NumPulmExacerbation"
)
y <- "Surv(time = tstart, time2 = tstop, event = as.numeric(CFRD) == 2)"
f <- as.formula(paste0(y, "~", paste0(xvars, collapse = "+")))
cox_res <- coxph(f, data = annual, cluster = eDWID)
# Forest plot and coefficients table
cox_res %>%
  tbl_regression(exponentiate = T)
ggcoef_table(cox_res,
  show_p_values = F, add_reference_rows = T, exponentiate = T,
  signif_stars = F, colour = F
)
```

@@

## 1 hour glucose alone

@@

ogtt <- read.csv("./1 hour OGTT in CFRD/Data_raw/ogtt.csv", na.strings = "")
pft <- read.csv("./1 hour OGTT in CFRD/Data_raw/pfts.csv", na.strings = "")
a1c <- read.csv("./1 hour OGTT in CFRD/Data_raw/a1c.csv", na.strings = "")
demo <- read.csv("./1 hour OGTT in CFRD/Data_raw/demographics.csv", na.strings = "")
dob <- read.csv("./1 hour OGTT in CFRD/Data_raw/dob.csv", na.strings = "")

@@

# read in date of modulator use
mod <- read.csv("./1 hour OGTT in CFRD/Data_raw/modulators.csv", na.strings = c("", " ", NA))
mod$Kalydeco.Start.Date <- as.Date(mod$Kalydeco.Start.Date, format = "%m/%d/%Y")
mod$Orkambi.Start.Date <- as.Date(mod$Orkambi.Start.Date, format = "%m/%d/%Y")
mod$Symdeko.Start.Date <- as.Date(mod$Symdeko.Start.Date, format = "%m/%d/%Y")
mod$Trikafta.Start.Date <- as.Date(mod$Trikafta.Start.Date, format = "%m/%d/%Y")
mod$modstart <- as.Date(NA)
mod$he_modstart <- as.Date(NA)
mod$me_modstart <- as.Date(NA)
for (i in 1:nrow(mod)) {
  mod[i, ]$modstart <-
    as.Date(pmin(mod[i, ]$Kalydeco.Start.Date, mod[i, ]$Orkambi.Start.Date, mod[i, ]$Symdeko.Start.Date, mod[i, ]$Trikafta.Start.Date, na.rm = T))
  mod[i, ]$he_modstart <- as.Date(pmin(mod[i, ]$Kalydeco.Start.Date, mod[i, ]$Trikafta.Start.Date, na.rm = T))
  mod[i, ]$me_modstart <- as.Date(pmin(mod[i, ]$Orkambi.Start.Date, mod[i, ]$Symdeko.Start.Date, na.rm = T))
}
modkeep <- mod[, c("MRN", "modstart", "he_modstart", "me_modstart")]
alldata <- merge(alldata, modkeep, by = "MRN", all.x = TRUE, all.y = FALSE)
alldata$onmod <- ifelse(!is.na(alldata$modstart) & alldata$modstart <= alldata$Date, 1,
  ifelse(!is.na(alldata$modstart) & alldata$modstart > alldata$Date, 0, NA)
)
alldata$onmod <- ifelse(is.na(alldata$modstart) & is.na(alldata$onmod), 0, alldata$onmod)

@@

##############
# COX MODELS #
##############
baseline_mult$Race <- factor(baseline_mult$Race, levels = c("Caucasian", "Black", "Other"))
# CPH model for baseline 1 hour glucose
f <- Surv(Follow_up_time_yrs, CFRD) ~ age + Sex + BMI + FEV1.per.pred + Baseline_1hr_glucose
cox1hr <- coxph(f, data = baseline_mult)
# Forest plot and coefficients table
cox1hr %>%
  tbl_regression(exponentiate = T)
ggcoef_table(cox1hr,
  show_p_values = F, add_reference_rows = T, exponentiate = T,
  signif_stars = F, colour = F
)
```

## 1 hour glucose and HbA1c

```{r}
cox1hra1c <- coxph(update(f, . ~ . + A1C), data = baseline_mult)
# Forest plot and coefficients table
cox1hra1c %>%
  tbl_regression(exponentiate = T)
ggcoef_table(cox1hra1c,
  show_p_values = F, add_reference_rows = T, exponentiate = T,
  signif_stars = F, colour = F
)
```

# Notes

- There are no participants over 31 in Christine's 1 hour OGTT analysis data, so I used age as a continuous variable.
- Because these models include age and sex, I also used raw BMI value as a continuous variable.
- Race was removed from the models because the group sizes are so small in the one hour dataset (only 3 Black participants).
- Pancreatic status was not available in the one hour glucose dataset, so was removed from the models.
